---
title: "Designing four-species subcommunities"
author: "Shane Hogle"
date: today
link-citations: true
---

# Setup

## Libraries
```{r}
#| output: false
library(here)
library(tidyverse)
library(ggplate)
```

## Global variables

```{r}
data <- here::here("data", "communities", "experiment_design")

# make processed data directory if it doesn't exist
fs::dir_create(data)
```

## Functions and vars

Species color vector
```{r}
my_colors <- c(
  "ANC_0403_10" = "#ffaaaa", "ANC_0403_70" = "#aa0000", "ANC_0403_80" = "#aa0000", "ANC_0403_90" = "#aa0000", 
  "ANC_1287_10" = "#ffeeaa", "ANC_1287_70" = "#d4aa00", "ANC_1287_80" = "#d4aa00", "ANC_1287_90" = "#d4aa00", 
  "ANC_1896_10" = "#ccffaa", "ANC_1896_70" = "#44aa00", "ANC_1896_80" = "#44aa00", "ANC_1896_90" = "#44aa00", 
  "ANC_1977_10" = "#aaccff", "ANC_1977_70" = "#0055d4", "ANC_1977_80" = "#0055d4", "ANC_1977_90" = "#0055d4",
  "EVO_0403_10" = "#ffaaee", "EVO_0403_70" = "#ff00cc", "EVO_0403_80" = "#ff00cc", "EVO_0403_90" = "#ff00cc", 
  "EVO_1287_10" = "#ffccaa", "EVO_1287_70" = "#ff6600", "EVO_1287_80" = "#ff6600", "EVO_1287_90" = "#ff6600", 
  "EVO_1896_10" = "#aaffee", "EVO_1896_70" = "#00ffcc", "EVO_1896_80" = "#00ffcc", "EVO_1896_90" = "#00ffcc", 
  "EVO_1977_10" = "#ccaaff", "EVO_1977_70" = "#7f2aff", "EVO_1977_80" = "#7f2aff", "EVO_1977_90" = "#7f2aff"
  )
```

For plotting plates
```{r}
plate_plot_steps <- function(df, well_colmun, val_column, step, colr_vec){
  ggplate::plate_plot(df,
  position = {{ well_colmun }},
  value = {{ val_column }},
  colour = colr_vec,
  plate_size = 96,
  plate_type = "round"
) + ggtitle(step)
}
```

# Construct quartet combinations

Get all possible quartets while excluding combinations of ANC/EVO of the same species

```{r}
combos04 <- t(combn(c("ANC_0403", "EVO_0403", "ANC_1287", "EVO_1287", "ANC_1896", "EVO_1896", "ANC_1977", "EVO_1977"),  4))
combos04 <- tibble("a" = combos04[, 1], "b" = combos04[, 2], "c" = combos04[, 3], "d" = combos04[, 4])
combos04_filt <- combos04 %>%
  filter(str_extract(a, "\\d+") != str_extract(b, "\\d+")) %>% 
  filter(str_extract(a, "\\d+") != str_extract(c, "\\d+")) %>% 
  filter(str_extract(a, "\\d+") != str_extract(d, "\\d+")) %>% 
  filter(str_extract(b, "\\d+") != str_extract(c, "\\d+")) %>%
  filter(str_extract(b, "\\d+") != str_extract(d, "\\d+")) %>% 
  filter(str_extract(c, "\\d+") != str_extract(d, "\\d+"))
```

Make plate layouts for the 4-species combos

```{r}
combos04_filt_wells <- combos04_filt %>% 
  slice(rep(1:n(), each = 4)) %>%
  mutate(a = paste(a, c(70, 10, 10, 10), sep = "_"),
         b = paste(b, c(10, 70, 10, 10), sep = "_"),
         c = paste(c, c(10, 10, 70, 10), sep = "_"),
         d = paste(d, c(10, 10, 10, 70), sep = "_")) %>% 
  add_count(a, name = "a_count") %>% 
  add_count(b, name = "b_count") %>% 
  add_count(c, name = "c_count") %>% 
  add_count(d, name = "d_count") %>% 
  arrange(desc(a_count), a, desc(b_count), b, desc(c_count), c, desc(d_count), d) %>%
  mutate(row = c(rep(LETTERS[c(5, 3, 6, 4)], each = 12), rep(LETTERS[1:2], each = 8))) %>% 
  group_by(row) %>% 
  mutate(col = if_else(!(row %in% c("A", "B")), str_pad(1:length(row), 2, pad = "0"), str_pad(1:length(row)+4, 2, pad = "0"))) %>% 
  ungroup() %>%
  mutate(well = paste0(row, col))
```

## Format and save

```{r}
combos04_filt_wells %>%
  group_by(a, b, c, d) %>% 
  mutate(microcosm_id = cur_group_id()) %>% 
  ungroup() %>% 
  mutate(a_sp = paste0(str_split_i(a, "_", 2), stringr::str_extract(str_split_i(a, "_", 1), "^.{1}")),
         b_sp = paste0(str_split_i(b, "_", 2), stringr::str_extract(str_split_i(b, "_", 1), "^.{1}")),
         c_sp = paste0(str_split_i(c, "_", 2), stringr::str_extract(str_split_i(c, "_", 1), "^.{1}")),
         d_sp = paste0(str_split_i(d, "_", 2), stringr::str_extract(str_split_i(d, "_", 1), "^.{1}")),
         a_f = str_split_i(a, "_", 3),
         b_f = str_split_i(b, "_", 3),
         c_f = str_split_i(c, "_", 3),
         d_f = str_split_i(d, "_", 3)) %>% 
  arrange(well) %>% 
  dplyr::select(microcosm_id, well, a, a_sp, a_f, b, b_sp, b_f, c, c_sp, c_f, d, d_sp, d_f) %>% 
  readr::write_tsv(here::here(data, "quartets_sample_composition_wide.tsv"))
```

```{r}
combos04_filt_wells %>%
  group_by(a, b, c, d) %>% 
  mutate(microcosm_id = cur_group_id()) %>% 
  ungroup() %>% 
  dplyr::select(a:d, well, microcosm_id) %>% 
  tidyr::pivot_longer(c(-well, -microcosm_id)) %>% 
  tidyr::separate(value, c("evo_hist", "strainID", "target_f")) %>% 
  dplyr::mutate(evo_hist = stringr::str_to_lower(evo_hist),
                strainID = paste0("HAMBI_", strainID),
                target_f = as.numeric(target_f)/100,
                n_species = 4) %>% 
  dplyr::select(-name) %>% 
  dplyr::relocate(microcosm_id, n_species) %>% 
  readr::write_tsv(here::here(data, "quartets_sample_composition_long.tsv"))
```

# Pipetting

Proceed in the order of steps below to construct master plates used to inoculate different conditions

## First pipetting step

::: {#fig-01}
```{r}
#| fig.width: 8
#| fig.height: 5
#| echo: false
plate_plot_steps(combos04_filt_wells, well, a, "Quartet Step 1", my_colors)
```
Layout for the first pipetting step for constructing quartets Different colors show different species/evolution combinations. The darkness of the color indicates whether a species/evolution combo is added at 70% (e.g., `_70`) or 10% (e.g., `_10`).
:::

## Second pipetting step

::: {#fig-02}
```{r}
#| fig.width: 8
#| fig.height: 5
#| echo: false
plate_plot_steps(combos04_filt_wells, well, b, "Quartet Step 2", my_colors)
```
Layout for the second pipetting step for constructing quartets Different colors show different species/evolution combinations. The darkness of the color indicates whether a species/evolution combo is added at 70% (e.g., `_70`) or 10% (e.g., `_10`).
:::

## Third pipetting step

::: {#fig-03}
```{r}
#| fig.width: 8
#| fig.height: 5
#| echo: false
plate_plot_steps(combos04_filt_wells, well, c, "Quartet Step 3", my_colors)
```
Layout for the third pipetting step for constructing quartets Different colors show different species/evolution combinations. The darkness of the color indicates whether a species/evolution combo is added at 70% (e.g., `_70`) or 10% (e.g., `_10`).
:::

## Fourth pipetting step

::: {#fig-04}
```{r}
#| fig.width: 8
#| fig.height: 5
#| echo: false
plate_plot_steps(combos04_filt_wells, well, d, "Quartet Step 4", my_colors)
```
Layout for the fourth pipetting step for constructing quartets Different colors show different species/evolution combinations. The darkness of the color indicates whether a species/evolution combo is added at 70% (e.g., `_70`) or 10% (e.g., `_10`).
:::