---
title: "Analysis of pairwise competition"
author: "Shane Hogle"
date: today
link-citations: true
abstract: "Here we analyze the community outpcomes from all competing species pairs."
---

# Introduction

Contains results from pairs of all streptomycin concentrations and trios for 0 streptomycin

# Setup

## Libraries

```{r}
#| output: false
library(tidyverse)
library(here)
library(fs)
library(scales)
library(patchwork)
library(broom)
library(ggraph)
library(tidygraph)

source(here::here("R", "communities", "community_composition", "utils_community_composition.R"))
```

## Global variables

```{r}
data <- here::here("data", "communities")
# make processed data directory if it doesn't exist
fs::dir_create(data)
```

# Read data

## Species abundances

```{r}
#| output: false
samp_pairs_9010 <- readr::read_tsv(here::here(data, "2sps_compiled_9010.tsv")) %>% 
  dplyr::rename(f = f_thresh)

samp_pairs_5050 <- readr::read_tsv(here::here(data, "2sps_compiled_5050.tsv")) %>% 
  dplyr::rename(f = f_thresh)
```

# Formatting

## Transfer 0 (starting proportions)

First get abundances from transfer 0 (masterplate)

```{r}
samp_pairs_fmt_mp_9010 <- samp_pairs_9010 %>% 
  dplyr::filter(str_detect(community_type, "master")) %>%
  format_master()
  
samp_pairs_fmt_mp_5050 <- samp_pairs_5050 %>% 
  dplyr::filter(str_detect(community_type, "master")) %>%
  format_master() %>% 
  mutate(community_id = paste0("P", as.numeric(str_extract(community_id, "\\d+"))+48))
```

## Transfer 8 (ending proportions)

Format abundances from the experiment, summarizing over replicates. Here we calculate the median frequency across biological replicates using `Hmisc::smedian.hilow` which computes the sample median and the outer quantiles (0.025 and 0.975).

```{r}
samp_pairs_fmt_exp_9010 <- samp_pairs_9010 %>% 
  dplyr::filter(community_type == "experiment") %>% 
  # make a combined evolution and species identifier and extract the community ID
  dplyr::mutate(sp = paste(str_to_upper(evo_hist), str_extract(strainID, "\\d+"), sep = "_"),
         community_id = str_extract(sample, "P\\d\\d")) %>%
  # calculate median and 95% CI across replicates
  dplyr::summarize(ggplot2::median_hilow(f), .by = c("community_id", "sp", "strep_conc")) %>% 
  # rename the y columns as f for compatibility
  dplyr::rename_with(.cols = starts_with("y"), \(x) str_replace(x, "y", "f")) %>% 
  # this step is important to ensure that when dfs are pivoted wider the 
  # sp_1 and sp_2 stay consistent 
  dplyr::arrange(community_id, sp, strep_conc) %>%
  # this creates an index for each species present in each community, it is needed
  # for the pivot to be consistent between the master plate and the samples
  dplyr::group_by(community_id, strep_conc) %>% 
  dplyr::mutate(n = 1:n()) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(id_cols = c(community_id, strep_conc), values_from = c(sp, f, fmin, fmax), names_from = n) %>% 
  dplyr::mutate(transfer = 8) %>% 
  arrange(sp_1, sp_2, strep_conc)
```

```{r}
samp_pairs_fmt_exp_5050 <- samp_pairs_5050 %>% 
  dplyr::filter(community_type == "experiment")  %>% 
  # make a combined evolution and species identifier and extract the community ID
  dplyr::mutate(sp = paste(str_to_upper(evo_hist), str_extract(strainID, "\\d+"), sep = "_"),
         community_id = str_extract(sample, "P\\d\\d")) %>%
  arrange(community_id, strep_conc) %>% 
  # there is only one replicate so we can't calculate median/mean
  dplyr::arrange(community_id, sp, strep_conc) %>%
  # this creates an index for each species present in each community, it is needed
  # for the pivot to be consistent between the master plate and the samples
  dplyr::group_by(community_id, strep_conc) %>% 
  dplyr::mutate(n = 1:n()) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(id_cols = c(community_id, strep_conc), values_from = c(sp, f), names_from = n) %>% 
  dplyr::mutate(transfer = 8) %>% 
  arrange(sp_1, sp_2, strep_conc) %>% 
  arrange(community_id) %>% 
  mutate(community_id = paste0("P", as.numeric(str_extract(community_id, "\\d+"))+48))
```

# Define competition outcomes

## Binomial sampling and Wilcox test

First need to determine which samples significantly decreased/increased from T0 to T8. We don't have enough biological replicates for to compute a statistic across replicates for [outcome variability](https://www.pnas.org/doi/10.1073/pnas.2302491120). However, we can estimate the mean fraction of A and quantify the inferential uncertainty of the mean by bootstrap resampling. We the used median proportion of species A from two biological replicates of each T8 pair as the probability of success (i.e., drawing species A) from 1000 draws (i.e., sequencing reads) from the binomial distribution. To determine whether the frequency of Species A significantly changed from T0 to T8, the means of the 1000 binomial draws for T0 and T8 were compared using a Wilcoxon rank sum test (N = 2000). Tests with Bonferroni multiple test corrected p values < 1e-5 were considered to represent significantly different T0 and T8 samples.  

run the tests

```{r}
set.seed(124341)

wc_test_9010 <- wc_test(samp_pairs_fmt_exp_9010, samp_pairs_fmt_mp_9010)
wc_test_5050 <- wc_test(samp_pairs_fmt_exp_5050, samp_pairs_fmt_mp_5050)
```

## Competition outcome rules

Here we set up the rules for defining the competition outcomes. In the pairwise competition experiments each species and evolutionary form was competed against all other species and evolutionary forms (excluding evolutionary forms of the same species because these cannot be resolved using amplicon data). Each competitor in a pair $\{A, B\}$ was allowed to invade from rare ($f^{R} = 0.1$) while the other competitor was common ($f^{C} = 0.9$) in duplicate for 8 growth cycles. Additionally we completed one experiment where each strain in the pair was started at roughly equal density. This resulted in 6 experiments for each competing pair. Using the abundance outcomes from both replicates of each the pairwise competition experiments for each pair we classified outcomes as "exclusion", "towards exclusion", "stable coexistence", "coexistence without evidence for mutual invasion", or "inconclusive" according to the following criteria (also see: https://www.science.org/doi/10.1126/science.adg0727)

### Exclusion

Requires one species ($A$) in a pair to be excluded ($f_{A} = 0$) in all 6 experiments for the species pair.

### Towards Exclusion

Relaxes the condition that ($f_{A} = 0$), but requires one species in a pair to have a significant decrease ($\Delta f_{8-0} < 0$) (Wilcoxon Test) in all 6 experiments for the species pair.

### Stable Coexistence
Both species $A$ and $B$ can stably coexist if each species can invade the other from low frequency $A_{f_{A} < f_{B}} \rightarrow B \vee B_{f_{B} < f_{A}} \rightarrow A$. Assignment of stable coexistence was made by comparing $sgn(\Delta f_{8-0}) = sgn(\bar{f_{8}} - f_{0})$ where $\bar{f_{8}}$ is the mean frequency of the species at the final sampling timepoint across all 6 experiments.

### Coexistence without evidence for mutual invasion
This outcome was assigned when each species had a non-zero frequency $f_{A|B} > 0$ in all 6 experiments at the final time point. 

### Inconclusive

This outcome was assigned when none of the above conditions were met.

```{r}
outcomes_classified <- bind_rows(wc_test_9010, wc_test_5050) %>% 
  arrange(sp_1, sp_2, strep_conc) %>% 
  group_by(sp_1, sp_2, strep_conc) %>%
  mutate(outcome = case_when(
    # condition 1
    sum(f_1_8 == 0) == n() | sum(f_1_8 == 1) == n() ~ "exclusion",
    # condition 2
    sum(change) == n() | sum(change) == -n() ~ "towards exclusion",
    # condition 3
    sum(sign(f_1_8 - f_1_0) == sign(mean(f_1_8)-f_1_0)) == n() ~ "stable coexistence",
    # condition 4
    sum(f_1_8 > 0) == n() ~ "coexistence no mutual invasion",
    # all others are inconclusive
    TRUE ~ "inconclusive"
  )) %>% 
  dplyr::ungroup()
```

EVO_1287 and EVO_1977 at 256 ug/ml streptomycin technically fullfill the requirements for the outcome "towards exclusion." However, the patterns at 64 and 16 ug/ml suggest that it could also be "coexistence no mutual invasion." Assigning coexistence no mutual invasion would break an intransitive loop apparent in the pairs network.

```{r}
outcomes_classified <- outcomes_classified %>% 
  mutate(outcome = if_else(sp_1 == "EVO_1287" & sp_2 == "EVO_1977" & strep_conc == 256, "coexistence no mutual invasion", outcome))

# save for later assessment of pairwise assembly rule
outcomes_classified %>% 
  mutate(coexists = if_else(str_detect(outcome, "coexistence"), 1, 0)) %>% 
  dplyr::select(strep_conc, sp_1, sp_2, coexists) %>% 
  distinct() %>% 
  write_tsv(here::here(data, "2sp_coexistence.tsv"))
```

## Plotting pairwise outpcomes

Construct final dataframe to be used for plotting

```{r}
samp_pairs_fmt <- dplyr::bind_rows(samp_pairs_fmt_mp_9010, samp_pairs_fmt_mp_5050, samp_pairs_fmt_exp_9010, samp_pairs_fmt_exp_5050) %>%
  dplyr::mutate(
    group = interaction(community_id, strep_conc),
    evo_group = dplyr::case_when(
      dplyr::if_all(c(sp_1, sp_2), \(x) stringr::str_detect(x, "ANC")) ~ "both_anc",
      dplyr::if_all(c(sp_1, sp_2), \(x) stringr::str_detect(x, "EVO")) ~ "both_evo",
      TRUE ~ "mix"
    )
  ) %>%
  dplyr::left_join(outcomes_classified,
                   by = dplyr::join_by(sp_1, sp_2, strep_conc),
                   relationship = "many-to-many") %>%
  dplyr::mutate(
    outcome = factor(outcome, levels = c("exclusion", 
                                         "towards exclusion", 
                                         "stable coexistence", 
                                         "coexistence no mutual invasion", 
                                         "inconclusive"))
  ) 
```

Create different lists of plots for the mixed (i.e. evo competed against anc) conditions

```{r}
samp_pairs_fmt_plots_split_a <- samp_pairs_fmt %>% 
  dplyr::filter(evo_group != "mix") %>% 
  dplyr::group_by(strep_conc, evo_group) %>% 
  dplyr::group_split() %>% 
  purrr::map(pair_plot)

samp_pairs_fmt_plots_split_b <- samp_pairs_fmt %>% 
  dplyr::filter(evo_group == "mix") %>% 
  dplyr::group_by(strep_conc) %>% 
  dplyr::group_split() %>% 
  purrr::map(pair_plot)
```

```{r}
#| warning: false
fig01 <- patchwork::wrap_plots(samp_pairs_fmt_plots_split_a, ncol = 2) +
  patchwork::plot_annotation(tag_levels = "A")
```

```{r}
#| eval: false
#| include: true
ggsave(
  here::here("figs", "coexistence_pairs.svg"),
  fig01,
  width = 7,
  height = 12,
  units = "in",
  device = "svg"
)
```

::: {#fig-01}
```{r}
#| fig-width: 7
#| fig-height: 14
#| warning: false
#| echo: false
fig01
```
Outcomes from pairwise cocultures of both ancestral (left, A:G) and both evolved (right, B:H) 0403, 1287, 1896 and 1977 species. Rows in the grid represent different streptomycin concentrations applied (A:B = 0 µg/ml, C:D = 16 µg/ml, E:F = 64 µg/ml, G:H = 256 µg/ml). Using the rules defined above for competitive outcomes, red lines show "exclusion," dark red show "towards exclusion," light blue lines show "stable coexistence," dark blue lines show "coexistence without evidence for mutual invasion," and green lines show "inconclusive" outcomes. Some statistical noise has been applied to point positions to prevent overlaps in the plot and aid in visualization.
:::

**NOTE:** in panel E (64 µg/ml streptomycin, ANC_0403 vs ANC_1977) both replicates survived with OD ~ 0.15 when ANC_0403 was started from 90% abundance but not when it was started at 10% abundance (OD ~ 0.04). In all replicates of 90%/10% starting abundance ANC_1977 was driven extinct.

```{r}
#| warning: false
fig02 <- patchwork::wrap_plots(samp_pairs_fmt_plots_split_b, ncol = 2) +
  patchwork::plot_annotation(tag_levels = "A")
```

```{r}
#| eval: false
#| include: true
ggsave(
  here::here("figs", "coexistence_pairs_mixed_hist.svg"),
  fig02,
  width = 7,
  height = 7,
  units = "in",
  device = "svg"
)
```

::: {#fig-02}
```{r}
#| fig-width: 7
#| fig-height: 7
#| warning: false
#| echo: false
fig02
```
Results from co-cultures of mixed ancestral and evolved combinations. Line colors and types are as in @fig-01.
:::

# Network

Here we will plot the pairwise competition outcomes as a network. **NOTE:** I don't have time/energy to figure out how to manage all the links in the networks so that they line up properly and don't overlap so there will need to be some postprocessing in inkscape to move some of the links so they don't overlap.

## Evolution and Streptomycin categories separate

Here we plot a separate graph for each streptomycin concentrations and also by different evolutionary groupings. For example, there is one graph for the competition outcomes of only ancestral species, there is one graph of the outcomes of only evolved species, and there is one graph for the outcomes of mixed competitions where an ancestral species competes against an evolved species.

```{r}
#| warning: false
nodes1 <- make_nodes(samp_pairs_fmt, evo_group, strep_conc)
edges1 <- make_edges(samp_pairs_fmt, evo_group, strep_conc)

graphs1 <- nest(nodes1, sps = -c(evo_group, strep_conc)) %>% 
  left_join(nest(edges1, pairs = -c(evo_group, strep_conc)),
            by = join_by(evo_group, strep_conc)) %>% 
  mutate(network = map2(sps, pairs, function(sps, pairs) tbl_graph(nodes = sps, edges = pairs, directed = T))) %>% 
  mutate(plot = map(network, function(network) plot_network_hierarchy(network, tune_angle = 1.5, n_rank = 7, n_break = 7)))
```


```{r}
#| warning: false
fig03 <- patchwork::wrap_plots(graphs1[[6]], nrow = 3, guides= "collect") +
  patchwork::plot_annotation(tag_levels = "A")
```

```{r}
#| eval: false
#| include: true
ggsave(
  here::here("figs", "coexistence_networks_nested_evo_strep.svg"),
  fig03,
  width = 8,
  height = 10,
  units = "in",
  device = "svg"
)
```

::: {#fig-03}
```{r}
#| fig-width: 8
#| fig-height: 10
#| warning: false
#| echo: false
fig03
```
Competitive hierarchy of species pairs separated by evolution grouping (only ancestral pairs, only evolved pairs, and mixed ancestral and evolved pairs) and streptomycin concentration. Subplots A-D are for only ancestral pairs, E-H for only evolved pairs, and I-L for mixed ancestral and evolved pairs. Subplots A, E, I show experiments under no streptomycin, B, F, J 16 µg/ml streptomycin, C, G, K for 64 µg/ml streptomycin, and D, H, L for 256 µg/ml streptomycin. For each evolution/streptomycin grouping, strains are rank ordered on the basis of the number of other strains they exclude, based on data shown in @fig-01 and @fig-02. Grey nodes represent strains (denoted by text), red arrows point from winning strain to losing strain, blue arrows indicate coexistence (ignore the arrow heads for blue, coulnd't figure out how to remove them for only a subset of the edges).
:::

## Only Streptomycin category separate

```{r}
#| warning: false
nodes2 <- make_nodes(samp_pairs_fmt, strep_conc)
edges2 <- make_edges(samp_pairs_fmt, strep_conc)

graphs2 <- nest(nodes2, sps = -c(strep_conc)) %>% 
  left_join(nest(edges2, pairs = -c(strep_conc)),
            by = join_by(strep_conc)) %>% 
  mutate(network = map2(sps, pairs, function(sps, pairs) tbl_graph(nodes = sps, edges = pairs, directed = T))) %>% 
  mutate(plot = map(network, function(network) plot_network_hierarchy(network, tune_angle = 1.5, n_rank = 7, n_break = 7)))
```

```{r}
#| warning: false
fig04 <- patchwork::wrap_plots(graphs2[[5]], nrow = 1, guides= "collect") +
  patchwork::plot_annotation(tag_levels = "A")
```

```{r}
#| eval: false
#| include: true
ggsave(
  here::here("figs", "coexistence_networks_nested_strep.svg"),
  fig04,
  width = 9,
  height = 4,
  units = "in",
  device = "svg"
)
```

::: {#fig-04}
```{r}
#| fig-width: 9
#| fig-height: 4
#| warning: false
#| echo: false
fig04
```
Competitive hierarchy of species pairs separated by streptomycin concentration. Subplot A shows experiments under no streptomycin, B with 16 µg/ml streptomycin, C with 64 µg/ml streptomycin, and D with 256 µg/ml streptomycin. For each evolution/streptomycin grouping, strains are rank ordered on the basis of the number of other strains they exclude, based on data shown in @fig-01 and @fig-02. Grey nodes represent strains (denoted by text), red arrows point from winning strain to losing strain, blue arrows indicate coexistence (ignore the arrow heads for blue, couldn't figure out how to remove them for only a subset of the edges)
:::
