---
title: "Formatting and plotting optical density data"
author: "Shane Hogle"
date: today
link-citations: true
---

# Introduction

In the main experiment, species pairs, trios, or quartets were serially passaged every 48 hours to fresh media containing the necessary concentration of streptomycin. The experiment was terminated after 8 of these growth cycles (16 days). The optical density (600 nm) was measured on alternating serial passage/growth cycles. This notebook contains results/plots for the optical density data. 

# Setup

## Libraries

```{r}
#| output: false
#| warning: false
#| error: false
library(tidyverse)
library(ggforce)
library(here)
library(fs)
library(scales)
library(ggh4x)
source(here::here("R", "utils_generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("_data_raw", "communities", "optical_density")
data <- here::here("data", "communities", "optical_density")

# make processed data directory if it doesn't exist
fs::dir_create(data)
```

## Read optical density data

```{r}
#| output: false
#| warning: false
#| error: false
pairs <- readr::read_tsv(here::here(data_raw, "20240606_pairs", "optical_density_formatted.tsv"))
tqs <- readr::read_tsv(here::here(data_raw, "20240829_tqs", "optical_density_formatted.tsv")) %>% 
  mutate(well = paste0(str_extract(well, "[A-H]"),
                               str_pad(str_extract(well, "\\d+"), 2, side = "left", pad = "0")))
```

## Read and format community information
```{r}
pairs_comp <- readr::read_tsv(here::here("data", "communities", "experiment_design", "pairs_sample_composition.tsv")) %>%
  dplyr::mutate(name = paste(str_to_upper(evo_hist), str_remove(strainID, "HAMBI_"), target_f*100, sep = "_")) %>% 
  dplyr::select(community_id, name) %>%
  dplyr::group_by(community_id) %>% 
  dplyr::mutate(id = 1:n()) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(names_from = id, values_from = name ) %>%
  dplyr::rename(a = `1`, b = `2`) %>% 
  mutate(a_sp = paste0(str_split_i(a, "_", 2), stringr::str_extract(str_split_i(a, "_", 1), "^.{1}")),
         b_sp = paste0(str_split_i(b, "_", 2), stringr::str_extract(str_split_i(b, "_", 1), "^.{1}")),
         a_f = str_split_i(a, "_", 3),
         b_f = str_split_i(b, "_", 3)) %>% 
  arrange(community_id) %>% 
  mutate(n_species = 2,
         sp = paste(a_sp, b_sp, sep = "|"),
         f = paste(a_f, b_f, sep = "|"))

trios_comp <- readr::read_tsv(here::here("data", "communities", "experiment_design", "trios_sample_composition_wide.tsv")) %>% 
  dplyr::mutate(n_species = 3,
                sp = paste(a_sp, b_sp, c_sp, sep = "|"),
                f = paste(a_f, b_f, c_f, sep = "|"))

quart_comp <- readr::read_tsv(here::here("data", "communities", "experiment_design", "quartets_sample_composition_wide.tsv")) %>% 
  mutate(n_species = 4,
         sp = paste(a_sp, b_sp, c_sp, d_sp, sep = "|"),
         f = paste(a_f, b_f, c_f, d_f, sep = "|"))
```

## Subset to pairs, trios, and quartets

Also we take the mean and std dev for the 2 replicates of each condition

```{r}
pairs_combo <- left_join(pairs_comp, pairs, by = join_by(community_id, n_species)) %>% 
  summarize(OD_mn = mean(OD),
            OD_sd = sd(OD),
            .by = c(sp, f, transfers, strep_conc)) %>% 
  mutate(strep_conc = as.factor(strep_conc))

trios_combo <- left_join(trios_comp, tqs, by = join_by(well, n_species)) %>% 
  summarize(OD_mn = mean(OD),
            OD_sd = sd(OD),
            .by = c(sp, f, transfers, strep_conc)) %>% 
  mutate(strep_conc = as.factor(strep_conc))

quart_combo <- left_join(quart_comp, tqs, by = join_by(well, n_species)) %>% 
  summarize(OD_mn = mean(OD),
            OD_sd = sd(OD),
            .by = c(sp, f, transfers, strep_conc)) %>% 
  mutate(strep_conc = as.factor(strep_conc))
```


# Plotting optical density

Here we will plot optical density over the batch growth cycles. For ease of visualization we exclude the first transfer because the OD was very high, and this happened in all conditions. We focus on the subsequent transfers because the OD has mostly stabilized after the first growth cycle. 

## Plotting function

```{r}
plot_od_grid <- function(df, remove_first_cycle = TRUE, ncol){
  pj <- ggplot2::position_jitterdodge(jitter.width=0.0,
                           jitter.height = 0.0,
                           dodge.width = 0.5,
                           seed=9)
  
  df %>% 
    dplyr::filter(if(remove_first_cycle) transfers > 2 else transfers > 0) %>%
    ggplot2::ggplot(aes(x = transfers, y = OD_mn, color = strep_conc, group = interaction(strep_conc, f))) + 
    ggplot2::geom_linerange(aes(ymin = OD_mn - OD_sd, ymax = OD_mn + OD_sd, color = strep_conc), position = pj) + 
    ggh4x::geom_pointpath(aes(shape = f), position = pj, mult = 0.2) +
    #ggplot2::geom_point() + 
    #ggplot2::geom_line(aes(linetype = f)) + 
    ggplot2::facet_wrap(~sp, ncol = ncol) +
    ggplot2::labs(x = "Growth cycle", y = "OD600", color = "Strep μg/ml", shape = "Mix ratio") +
    ggplot2::scale_color_viridis_d() +
    ggplot2::theme_bw() + 
    ggplot2::theme(strip.background = element_blank(),
        legend.position = "bottom", 
        axis.text = element_text(size = 8),
        strip.text = element_text(size = 8))
}
```

## Pairs

```{r}
#| warning: false
fig_pairs <- plot_od_grid(pairs_combo, remove_first_cycle = TRUE, ncol = 4)

# ggsave(
#   here::here("figs", ".svg"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "svg"
# )
# 
# ggsave(
#   here::here("figs", ".png"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "png"
# )
```

::: {#fig-01}
```{r}
#| fig-width: 7
#| fig-height: 11
#| warning: false
#| echo: false
fig_pairs
```
Species pairs optical density (y-axis) over series of 48 hour growth cycles (x-axis, 16 days total). Different panels in the grid contain distinct species pairs (A = ancestral, E = strep evolved). Colors show Streptomycin concentration in μg/ml and point shape shows the initial mixing ratio of each species (in the same order as specie listed in each panel). Note that data from the first growth cycle (after 48 hours) is excluded for ease of viewing.
:::

### Pairs with first growth cycle

::: {#fig-02}
```{r}
#| fig-width: 7
#| fig-height: 11
plot_od_grid(pairs_combo, remove_first_cycle = FALSE, ncol = 4)
```
As in @fig-01 but including data from the first growth cycle.
:::


## Trios

```{r}
#| warning: false
fig_trios <- plot_od_grid(trios_combo, remove_first_cycle = TRUE, ncol = 4)

# ggsave(
#   here::here("figs", ".svg"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "svg"
# )
# 
# ggsave(
#   here::here("figs", ".png"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "png"
# )
```

::: {#fig-03}
```{r}
#| fig-width: 7
#| fig-height: 14
#| warning: false
#| echo: false
fig_trios
```
Species trios optical density (y-axis) over series of 48 hour growth cycles (x-axis, 16 days total). Different panels in the grid contain distinct species pairs (A = ancestral, E = strep evolved). Colors show Streptomycin concentration in μg/ml and point shape shows the initial mixing ratio of each species (in the same order as species listed in each panel). Note that data from the first growth cycle (after 48 hours) is excluded for ease of viewing.
:::

### Trios with first growth cycle

::: {#fig-04}
```{r}
#| fig-width: 7
#| fig-height: 14
plot_od_grid(trios_combo, remove_first_cycle = FALSE, ncol = 4)
```
As in @fig-03 but including data from the first growth cycle.
:::

## Quartets

```{r}
#| warning: false
fig_quarts <- plot_od_grid(quart_combo, remove_first_cycle = TRUE, ncol = 4)

# ggsave(
#   here::here("figs", ".svg"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "svg"
# )
# 
# ggsave(
#   here::here("figs", ".png"),
#   fig01,
#   width = 7,
#   height = 12,
#   units = "in",
#   device = "png"
# )
```

::: {#fig-05}
```{r}
#| fig-width: 7
#| fig-height: 7.5
#| warning: false
#| echo: false
fig_quarts
```

Species quartets optical density (y-axis) over series of 48 hour growth cycles (x-axis, 16 days total). Different panels in the grid contain distinct species pairs (A = ancestral, E = strep evolved). Colors show Streptomycin concentration in μg/ml and point shape shows the initial mixing ratio of each species (in the same order as species listed in each panel). Note that data from the first growth cycle (after 48 hours) is excluded for ease of viewing.
:::

### Quartets with first growth cycle

::: {#fig-06}
```{r}
#| fig-width: 7
#| fig-height: 7.5
plot_od_grid(quart_combo, remove_first_cycle = FALSE, ncol = 4)
```
As in @fig-05 but including data from the first growth cycle.
:::
