---
title: "Formatting Rbec output from 2024-10-20 sequenced by BTK Turku using custom HAMBI Illumina v3 primers"
author: "Milla Similä & Shane Hogle"
date: today
link-citations: true
abstract: "Results from trios and quartets in 0 ug/ml streptomycin concentration and their masterplates from Milla's community assembly experiment. Also includes a couple pairs that failed last time that we redid in this sequencing run."
---

# Setup

## Libraries

```{r}
#| output: false
#| warning: false
#| error: false
library(tidyverse)
library(here)
library(fs)
library(archive)
library(scales)
source(here::here("R", "utils_generic.R"))
source(here::here("R", "communities", "amplicon", "utils_amplicon.R"))
```

## Global variables

```{r}
#| output: false
#| warning: false
#| error: false
data_raw <- here::here("_data_raw", "communities", "20241024_BTK_illumina_v3")
data <- here::here("data", "communities", "20241024_BTK_illumina_v3")
amplicontar <- here::here(data_raw, "rbec_output.tar.gz")

# make processed data directory if it doesn't exist
fs::dir_create(data)

# create temporary location to decompress
tmpdir <- fs::file_temp()
```

# Read metadata

```{r}
#| output: false
#| warning: false
#| error: false
mddf <- readr::read_tsv(here::here(data_raw, "20241024_BTK_illumina_v3_metadata.tsv")) %>% 
  mutate(sample = str_trim(sample))
spdf <- readr::read_tsv(here::here(data_raw, "sample_compositions.tsv")) %>% 
  mutate(sample = str_trim(sample))
```

# Read Rbec raw counts tables 

## Untar Rbec output tarball

```{r}
archive::archive_extract(
  amplicontar,
  dir = tmpdir,
  files = NULL,
  options = character(),
  strip_components = 0L
)
```

## Setup directory structure

```{r}
tabdir <- here::here(tmpdir, "rbec_output")
samppaths <- fs::dir_ls(tabdir)
sampnames <- path_split(samppaths) %>% 
  map_chr(dplyr::last)
```

## Read

```{r}
straintabs <- paste0(samppaths, "/strain_table.txt") %>% 
  set_names(sampnames) %>% 
  map_df(
  read_tsv,
  skip = 1,
  col_names = c("rRNA16S_locus","count"),
  show_col_types = FALSE, 
  .id = "sample")
```

## Clean up

```{r}
# remove decompressed directory from temp location
fs::dir_delete(tmpdir)
```

# Format

```{r}
straintabs_norm <- straintabs %>% 
  # Calls function that normalize counts by 16S copy number
  normalize_by_copy() %>% 
  # Calls function that completes all combinations of 23 species
  # this is important because some species go extinct that should be in the samples
  # and we need to have those in the final table
  complete_combos() %>% 
  # removes any extra whitespace on sample names
  mutate(sample = str_trim(sample))
```

Make final table

```{r}
# Later we will take advantage of the fact that for species not supposed to be
# in a sample the prior left_join will have filled the evo_hist category with an
# NA. We can then filter using this NA value
finaltable <- left_join(straintabs_norm, mddf, by = join_by(sample)) %>% 
  left_join(spdf, by = join_by(sample, strainID)) %>% 
  group_by(sample) %>% 
  dplyr::select(-genus, -species) %>% 
  mutate(f_raw = count_correct/sum(count_correct),
         sample_tot_raw = sum(count_correct)) %>% 
  ungroup() %>% 
  relocate(sample_tot_raw, f_raw, target_f, evo_hist, .after = count_correct) %>% 
  # reformats sample names for pairs so that they are same as in the the batch from
  # 20240711
  mutate(sample = case_when(n_species == 2 ~ notes, 
                            TRUE ~ sample))
```

# Analysis

Let's quickly compare and look the at number of reads from the different kinds of experiment categories. Negative controls are samples from the experiment that contained growth medium but no cells. We included them to make sure that there was not contamination between wells during the experiment. Positive controls contain a mixsture of all 23 species (not sure why this was done), and some wells are just empty but were processed/sequenced anyway for convenience of the person preparing the libraries. Experiment samples are those that came from the actual experiment, and masterplate samples are the defined mixtures we made to inoculate the experiment. We sequenced these masterplates so that we would have T0 information for each experimental treatment.

::: {#fig-01}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable %>% 
  dplyr::select(sample, sample_tot_raw, community_type) %>% 
  distinct() %>% 
  ggplot(aes(x = sample_tot_raw)) +
  geom_histogram(bins = 20, aes(fill = community_type)) +
  scale_x_log10() + 
  annotation_logticks(sides = "b", color="grey30") +
  labs(y = "Sample count", x = "Total reads per sample") +
  facet_wrap(~ community_type, scales="free_y", nrow = 2)
```
Frequency distribution of total read counts per sample in the experimental samples (red) and the negative controls (blue).
:::

The smallest number of reads in the experimental samples is `{r} min(finaltable[ grep("^exp", finaltable$community_type), ]$sample_tot_raw)` and the mean is `{r} format(mean(finaltable[ grep("^exp", finaltable$community_type), ]$sample_tot_raw), scientific=FALSE)` which is considerably more than the smallest number of reads `{r} min(finaltable[ grep("^neg", finaltable$community_type), ]$sample_tot_raw)` and the mean `{r} mean(finaltable[ grep("^neg", finaltable$community_type), ]$sample_tot_raw)` in the negative controls.

This looks really good. The number of reads in the negative controls are always very low (similar to "empty" samples and always an order of magnitude less than masterplates and experimental samples). The number of reads in the negative controls will depend on the sequencing depth used and in this batch compared to the first one with pairs we have used a higher sequencing depth (or less multiplexing) so that the average number of reads is higher here. 

## Negative controls

First, we’ll check whether the experimental and plate negative controls look good

Lets check what species are contaminating negative controls. Most appear to be low abundance contaminants of species excluded from the experiment. Probably this is cross talk from the positive controls and other samples

::: {#fig-02}
```{r}
#| fig.width: 8
#| fig.height: 7
finaltable %>% 
  filter(str_detect(community_type, "^neg")) %>% 
  filter(f_raw > 0) %>%
  contam_histogram(trans=TRUE, x = f_raw) +
  labs(x = "Frequency distribution of each species in the negative controls", y = "Sample count") +
  facet_wrap(~strainID)
```
Frequency distribution of all species in the negative controls
:::

::: {#fig-03}
```{r}
#| fig.width: 8
#| fig.height: 3
finaltable %>% 
  filter(str_detect(community_type, "^neg")) %>% 
  filter(str_detect(strainID, "0403|1287|1896|1977")) %>% 
  filter(f_raw > 0) %>% 
  contam_histogram(trans=TRUE, x = f_raw) +
  labs(x = "Frequency distribution of each species in the negative controls", y = "Sample count")
```
Frequency distribution of only target target species in the negative controls
:::

<!-- Again, I think this is really good. Negative controls are strange to sequence because they have such low biomass usually and then weird things can happen with the PCRs in the library preparation stages. However, let’s make sure negative control samples have low OD600 in the endpoint samples. This would indicate growth and that what we are seeing in the histograms is truly just a sequencing artifact -->

```{r}
#| output: false
#| warning: false
#| error: false
od <- read_tsv(here::here("_data_raw", "communities", "optical_density", "20240829_tqs", "optical_density_formatted_102024.txt")) %>% 
  mutate(well = stringr::str_replace(well,
                     pattern = "([0-9]+)", 
                     replacement = {\(x) stringr::str_pad(string = x, width = 2,  side = "left",  pad = "0")}))
```

All negative control have very low OD < 0.1 and it is safe to assume that they are clean.

```{r}
df <- left_join(finaltable, od, by = join_by(sample, transfers==transfer, replicate, strep_conc)) %>% 
  filter(str_detect(community_type, "^neg")) %>% 
  dplyr::select(sample, OD, well, community_type, n_species.x, n_species.y, replicate) %>% 
  distinct() %>% 
  filter(!is.na(OD))
```

All negative control have very low OD < 0.1 and low read counts it is safe to assume that they are clean.

::: {#fig-04}
```{r}
#| fig.width: 8
#| fig.height: 5
df %>% 
  left_join(finaltable, by = join_by(sample, community_type, replicate, n_species.x==n_species)) %>% 
  filter(f_raw > 0) %>% 
  filter(str_detect(strainID, "0403|1287|1896|1977")) %>% 
  dplyr::select(sample_tot_raw, sample, OD, community_type) %>% 
  distinct() %>% 
  pivot_longer(c(-sample, -community_type)) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 5) +
  facet_grid(~name, scales = "free_x")
```
Optical density (OD) and total read count distributions of negative controls.
:::

## Positive controls

The positive community controls should each have all 23 species.

```{r}
finaltable %>% 
  filter(str_detect(community_type, "^pos") & count_correct > 0) %>% 
  group_by(sample) %>%
  mutate(n_sp_detected = n()) %>% 
  distinct(sample, sample_tot_raw, n_sp_detected)
```
2/3 community controls have plenty of reads and look normal. One has under 1000 reads in total and it only contains the 4 focal species (0403, 1287, 1896, 1977). It looks to me that this could be a mislabeled negative control?

## Misassigned reads

These libraries were only prepared with samples containing HAMBI_0403, HAMBI_1287, HAMBI_1896, and HAMBI_1977 so any time species other than these show up is just an incorrect assignment by Rbec or index leakage/crosstalk from the positive community controls. Let's check quickly how many of these there are...

```{r}
finaltable %>% 
  # filter out community controls since they have all hambi species
  filter(!(sample %in% c("4_p09_A01", "4_p10_A01", "4m_p18_A01" ))) %>% 
  filter(!str_detect(strainID, "0403|1287|1896|1977")) %>% 
  filter(count_correct > 0) %>% 
  arrange(desc(count_correct))
```

there are only very small number of reads mapping to off-target species per experimental sample probably from crosstalk leakage.

## Samples with few total reads

Some of the experimental pairs had streptomycin concentrations higher than any of the species individually could tolerate. We would expect then that no species would grow successfully in these samples and that the overall biomass would be very low, thus resulting in a low number of recovered reads from these samples.

To look into this. first let’s check which samples have very low OD600 in the endpoint samples. We’ll filter out samples with an OD of less than 0.1 and also samples with fewer than 1000 reads. It is generally good practice to exclude samples with low number of reads.

```{r}
left_join(finaltable, od, 
          by = join_by(sample, transfers==transfer, replicate, strep_conc)) %>% 
  filter(transfers == 8) %>% 
  filter(OD < 0.1 | sample_tot_raw < 3000) %>% 
  distinct(sample, community_type, OD, sample_tot_raw) %>% 
  count(community_type)
```

Samples with low reads or low OD are all negative controls or the one positive control that failed (plus another positive control `4_p10_A01` but this must be from mislabeling because there were no positive controls actually on the experimental plates). 

## Filter to target species

Exclude positive and negative controls and also exclude species that are not in the focal 4

```{r}
finaltable_exp_mstr <- finaltable %>% 
  filter(community_type %nin% c("empty", "neg_control", "pos_control")) %>% 
  # no low-read samples to filter
  #filter(sample %nin% lowread_samps) %>% 
  filter(strainID %in% c("HAMBI_0403", "HAMBI_1287", "HAMBI_1896", "HAMBI_1977")) %>% 
  group_by(sample) %>% 
  mutate(f_raw_targetsp = count_correct/sum(count_correct),
         sample_tot_targetsp = sum(count_correct)) %>% 
  ungroup() %>% 
  relocate(c(sample_tot_targetsp, f_raw_targetsp), .after = f_raw)
```

## Masterplate samples

### Pairs

Something a bit weird is going on with two of the masterplate pairs. Samples P19 and P20 both somehow have ~20 of HAMBI_0403 present in them. However, the rest of the species composition looks like how we expect them. I think we will keep these two masterplate samples and just filter out the HAMBI_0403 that shouldn't be there

```{r}
finaltable_exp_mstr %>%
  filter(str_detect(community_type, "masterplate") & n_species == 2)  %>% 
  dplyr::select(sample, strainID, f_raw_targetsp, target_f, evo_hist) %>% 
  mutate(f_raw_targetsp = round(f_raw_targetsp, 2)) %>% 
  distinct()
```

::: {#fig-05}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>%
  filter(str_detect(community_type, "masterplate") & n_species == 2) %>% 
  contaminated_barplot(threshold = 0, quartet = FALSE, y=f_raw_targetsp) +
  ggtitle("Masterplate pairs with contamination")
```
Species composition of masterplate pairs that contain any percentage (> 0%) of a species that shouldn’t be there (shown in grey to black).
:::

### Trios

Trio masterplates look quite good. Yes, there are some samples where a species that shouldn't be there is > 3%, but these are always < 5% and there are no samples with major contamination 

::: {#fig-06}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>%
  filter(str_detect(notes, "trios") & str_detect(notes, "masterplate")) %>% 
  contaminated_barplot(threshold = 0.01, quartet = FALSE, y=f_raw_targetsp) +
  ggtitle("Masterplate trios with contamination")
```
Species composition of masterplate pairs that contain >= 1% of a species that shouldn’t be there (shown in grey to black).
:::

::: {#fig-07}
```{r}
#| fig.width: 8
#| fig.height: 4
finaltable_exp_mstr %>%
  filter(str_detect(notes, "trios") & str_detect(notes, "masterplate")) %>% 
  filter(is.na(evo_hist)) %>% 
  filter(f_raw > 0) %>% 
  contam_histogram(trans = FALSE, x = f_raw_targetsp) +
  labs(x = "Species frequency", y = "Sample count") +
  ggtitle("Frequency distribution of misassigned species in masterplate trios")
```
Frequency distribution of each species in masterplate samples where it **should not** occur. The 1% mark (commonly accepted as an acceptable Illumina cross-talk standard) is demarcated with a dashed line. 
:::

::: {#fig-08}
```{r}
#| fig.width: 8
#| fig.height: 3
finaltable_exp_mstr %>%
  filter(str_detect(notes, "trios") & str_detect(notes, "masterplate")) %>% 
  filter(!is.na(evo_hist)) %>% 
  filter(f_raw > 0) %>% 
  contam_histogram(trans = FALSE, x = f_raw_targetsp) +
  labs(x = "Species frequency", y = "Sample count") +
  ggtitle("Frequency distribution of focal species in masterplate trios")
```
Frequency distribution of each species in all samples where it **should** occur.
:::

### Quartets

One weakness of this contamination-detection approach is that there is no way to tell from this data whether a quartet masterplate is contaminated because every quartet (by definition) is inoculated with all of the 4 species at different frequencies and with different evolutionary histories. See @sec-quartets2 below for more information. However, we can still visualize the composition of those quartets and whether they seem to be close to what we expected to put in.

::: {#fig-09}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>%
  filter(str_detect(notes, "quartets") & str_detect(notes, "masterplate")) %>% 
  contaminated_barplot(threshold = 0.01, quartet = TRUE, y = f_raw_targetsp) +
  ggtitle("All masterplate quartets")
```
Species composition of masterplate quartets.
:::

These look really good and are clearly following the correct pattern/distribution for the quartets.

## Experimental samples

We may need to use a higher threshold than 1% (e.g., 3% frequency) because over 3% we can reliably say that a species is present and it is not due to index cross talk.

::: {#fig-10}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>% 
  filter(str_detect(community_type, "experiment")) %>% 
  filter(is.na(evo_hist)) %>% 
  filter(f_raw > 0.03 ) %>% 
  contam_histogram(trans=TRUE, x = f_raw_targetsp) +
  labs(x = "Species frequency", y = "Sample count") +
  ggtitle("Frequency distribution of each species exceeding the 3% threshold")
```
Frequency distribution of each species in experimental samples where it **should not** occur and where it exceeds a 3% threshold.
:::

```{r}
finaltable_exp_mstr %>% 
  # there are only experiment samples left
  filter(str_detect(community_type, "experiment")) %>% 
  filter(is.na(evo_hist)) %>% 
  filter(f_raw_targetsp > 0.03 ) %>% 
  dplyr::select(sample, strainID, f_raw_targetsp) %>% 
  arrange(desc(f_raw_targetsp))
```

```{r}
#| eval: true
#| echo: false
#| output: false
myfilter <- function(.data, myexpresion){
  .data %>% 
    filter(str_detect(community_type, "experiment")) %>% 
    filter(is.na(evo_hist)) %>% 
    filter( {{ myexpresion }} ) %>% 
    distinct(sample) %>% 
    nrow()
}

thresh <- 0.08
totsamp <- nrow(distinct(finaltable_exp_mstr, sample))
contamsamp <- myfilter(finaltable_exp_mstr, f_raw_targetsp >= 0.03)
contamsamp_lo <- myfilter(finaltable_exp_mstr, f_raw_targetsp >= 0.03 & f_raw_targetsp < thresh)
contamsamp_hi <- myfilter(finaltable_exp_mstr, f_raw_targetsp >= thresh)
```

There are `{r} contamsamp` samples (`{r} round(contamsamp/totsamp*100)`% of `{r} totsamp` samples) that exceed a 3% crosstalk threshold. `{r} contamsamp_lo` samples (`{r} round(contamsamp_lo/totsamp*100)`% of `{r} totsamp` samples) have contamination <= `{r} thresh*100`% and `{r} contamsamp_hi` samples (`{r} round(contamsamp_hi/totsamp*100)` % of `{r} totsamp` samples) have > `{r} thresh*100`% of a species that should not be there. In most cases the species in highly contaminated samples is HAMBI_1977. I think these `{r} contamsamp_hi` samples are probably at very high risk for legitimate contamination. It is likely that the rest of the `{r} contamsamp_lo` samples with <= `{r} thresh*100`%  contamination are not contaminated but just crosstalk outliers.

### Pairs

Here just taking a closer look at the pairs from the experiment that appear to be contaminated.

::: {#fig-11}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>% 
  filter(str_detect(community_type, "experiment") & n_species == 2) %>% 
  contaminated_barplot(threshold = 0.03, quartet = FALSE, y=f_raw_targetsp) +
  ggtitle("Experiment pairs with > 3% contamination")
```
Species composition of pairs masterplates samples that contain > 3% of a species that shouldn’t be there (shown in grey to black). 
:::

All experimental pair samples look pretty good. I think we should only exclude samples where the contamination is very high (over ~50% of the sample) but those with around 10% or less contaminant I think can be retained. So we will not need filter any of these out. 

### Trios

Here just taking a closer look at the trios from the experiment that appear to be contaminated.

::: {#fig-12}
```{r}
#| fig.width: 8
#| fig.height: 5
finaltable_exp_mstr %>% 
  filter(str_detect(community_type, "experiment") & n_species == 3) %>% 
  contaminated_barplot(threshold=0.03, quartet = FALSE, y=f_raw_targetsp) +
  ggtitle("Experiment trios with > 3% contamination")
```
Species composition of experiment trio samples that contain > 3% of a species that shouldn’t be there (shown in grey to black). 
:::

Only 2 trio samples look really bad (over 10 %, more like 50 % contamination). Again, I think we should only exclude samples where the contamination is very high (over ~50% of the sample) but those with around 10% or less contaminant I think can be retained, and I will discard the contaminating sequences.

```{r}
contamtrios <- finaltable_exp_mstr %>% 
  filter(str_detect(community_type, "experiment") & n_species == 3) %>% 
  filter(is.na(evo_hist)) %>% 
  filter(f_raw_targetsp > 0.13  ) %>% 
  pull(sample)
```

### Quartets {#sec-quartets2}

Again, one weakness of this approach is that there is no way to tell from this data whether a quartet is contaminated because every quartet (by definition) is inoculated with all of the 4 species at different frequencies and with different evolutionary histories. Thus we would expect to see either all 4 species present or a subset of the 4 species in the quartet samples, but we cannot tell if these patterns come from contamination or not. Moving forward we just have to assume that the quartets are OK... However, we've found so little contamination in our other experimental samples and all samples were handled in the experiment and processed for sequence library preparation in the same way. I think it is safe to assume that contamination in the quartets is minimal issue. 

::: {#fig-13}
```{r}
#| fig.width: 10
#| fig.height: 12
finaltable_exp_mstr %>% 
  filter(str_detect(community_type, "experiment")) %>% 
  filter(n_species == 4) %>% 
  group_by(sample) %>% 
  mutate(id = cur_group_id()) %>% 
  ungroup() %>% 
  mutate(facet = ntile(id, 4)) %>% 
  contaminated_barplot(threshold = 0, quartet = TRUE, y=f_raw_targetsp) +
  ggtitle("All experimental quartets") +
  facet_wrap(~facet, scales="free_x", nrow=4) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    legend.position = "bottom"
    )
```
Species composition of masterplate quartets.
:::

# Export

```{r}
finaltable_exp_mstr_export <- finaltable_exp_mstr %>% 
  # first remove samples with low read count (< 3000 reads)
  # filter(sample %nin% lowread_samps) %>% 
  # next remove masterplate pair samples that were highly contaminated
  # filter(sample %nin% contampairsmaster) %>% 
  # remove trio samples that were highly contaminated
  filter(sample %nin% contamtrios) %>% 
  # exclude any remaining counts from species that shouldnt be there using again
  # the fact that evo_hist should be NA for these species
  filter(!is.na(evo_hist)) %>% 
  # because we set 3% as our limit of detection we set read counts of species
  # less than 1% to 0
  mutate(count_correct_thresh = if_else(f_raw_targetsp <= 0.03, 0, count_correct)) %>% 
  # now calculate a new relative abundance based only on the species that should
  # be present and that are > 3% relative abundance
  group_by(sample) %>% 
  mutate(f_thresh = count_correct_thresh/sum(count_correct_thresh)) %>% 
  ungroup() %>%
  # for marking which sequencing batch these came from 
  mutate(batch = "run202410124") %>% 
  dplyr::select(sample, strainID, evo_hist, count_correct_thresh, f_thresh,
                target_f_masterplate = target_f, replicate, strep_conc, 
                transfers, n_species, community_type, plate_well, batch)
```

## Pairs

Write the pair samples

```{r}
finaltable_exp_mstr_export %>% 
  filter(n_species == 2) %>% 
  arrange(sample) %>% 
  write_tsv(here::here(data, "pairs_counts.tsv"))
```

## Trios

Write the trio samples

```{r}
finaltable_exp_mstr_export %>% 
  filter(n_species == 3) %>% 
  arrange(sample) %>% 
  write_tsv(here::here(data, "trios_counts.tsv"))
```

## Quartets

Write the quartet samples

```{r}
finaltable_exp_mstr_export %>% 
  filter(n_species == 4) %>% 
  arrange(sample) %>% 
  write_tsv(here::here(data, "quartets_counts.tsv"))
```
